\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[a4paper, total={8in, 10.5in}]{geometry}
\usepackage{rotating}
\usepackage{afterpage}
\usepackage{url}




\title{Godanov Scheme}
\author{Maximilian Williams}
\date{June 2021}

\begin{document}

\maketitle

\section*{What is this?}
A little discussion on the Godanov scheme

\section*{Problem}
I was trying to find an explicit scheme for solving the advection-diffusion equation. The diffusion part is not an issue, so lets just look at the advection equation.
\begin{equation}
	\frac{\partial T}{\partial t} + v \frac{\partial T}{\partial x} = 0
\end{equation}
We will employ the goodanove scheme to numerically solve this equation. We discritize $\frac{\partial T}{\partial t}$ as:
\begin{equation}
	\frac{\partial T}{\partial t} = \frac{T^{n+1}_{i} - T^{n}_{i}}{\Delta t}.
\end{equation}
Now here comes the trick with the Godanov scheme. If $v>0$, then we use a upstream scheme:
\begin{equation}
	\frac{\partial T}{\partial x} = \frac{T^{n}_{i-1} - T^{n}_{i}}{\Delta x}.
\end{equation}
while if $v<0$ we use a downwind scheme:
\begin{equation}
	\frac{\partial T}{\partial x} = \frac{T^{n}_{i+1} - T^{n}_{i}}{\Delta x}.
\end{equation}
In doing this, we never use information that travelled against the advection velocity $v$. The MEPDE for this scheme looks like:
\begin{equation}
	\frac{\partial T}{\partial t} + v \frac{\partial T}{\partial x} = \mid v \mid \frac{\Delta x}{2} (1 - \mid v \mid  \frac{\Delta t}{\Delta x}) \frac{\partial^2 T}{\partial x^2} + \mathcal{O}({\Delta x}^2, {\Delta t}^2 ).
\end{equation}
So, this scheme has some numerical diffusion, with coeffecient $\kappa' = \mid v_c \mid \frac{\Delta x}{2} (1 - \mid v_c \mid) $, where we will use $v_c$ to denote the maximum velocity in our problem. Lets overeastimate this a little so that $\kappa' = \mid v_c \mid \frac{\Delta x}{2}$. We have neglected the diffusion term in our advection diffusion equation, but lets add it in to see how this numerical diffusion affects our numerical solution:
\begin{equation}
	\frac{\partial T}{\partial t} + v \frac{\partial T}{\partial x} = (\kappa + \kappa') \frac{\partial^2 T}{\partial x^2},
\end{equation}
So long as $\kappa >> \kappa'$ this numerical diffusivity is negligable and this scheme is accurate. However, this may require a very fine grid and so take lots of memory.









\end{document}
